Union and intersection contracts: theory vs practice

### 2021-04-06 paper.tex
WIP: reduce text

The explorative work presented in this paper is motived by the design of the
Nickel language\cite{NickelRepo}. Nickel is a configuration programming
language, meaning that a Nickel program evaluates to a value that must be
straightforwardly serializable to JSON, YAML or alike.  Nickel aims at being
lightweight, easy to write and easy to embed in a larger system, to serve either
directly as a native language to specify configuration, or as a meta-language to
generate configuration. The focus is first and foremost practical: Nickel
originated as a rethink of the language of the Nix package manager, used to
describe one of the biggest sets of software packages ever built\cite{repology}.
Nickel targets build systems and cloud deployment tools (Kubernetes, Terraform
or NixOps) as well. While there's no hesitation in incorporating recent research
when it serves a well-defined purpose, the design decisions are guided by the
industrial use cases.

@@

Our aim is that writing simple Nickel should not be more complex than writing
JSON or YAML. Unlike JSON, though, Nickel anticipates large configurations by
being both programmable and typed. In the configuration setting, there is a
singular dilemma with respect to typing: since a program is run on fixed inputs
and is expected to terminate, any relevant type error should show up anyway.
Why bother with the complexity of a static type system? On the other hand, more
and more software systems offload complexity to configurations, such as the
infrastructure as code paradigm used in cloud deployment. When complexity grows,
static types becomes attractive again. For reusable code — that is, library
functions — static types brings all the usual widely acknowledged benefits.

This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}
that mixes both static typing and dynamic typing. Unlike gradually-typed
languages like TypeScript, we do not seek to build a complex type system that
tries hard to accept most of the idioms naturally arising in dynamically typed
code. Rather, we choose to provide a reasonably expressive type system with good
inference properties, that make typing generic function operating on data easy ,
but may require to resort to untyped code for more exotic expressions.  In any
case, this is not a surrender, as Nickel provides a complementary mechanism for
more advanced data validation: contracts.

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is crucial to good error reporting. It turns out these contracts are closely
related to the type casts introduced by gradual typing, modulo blame: both
\cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the value of contracts
as a safe interface between typed and untyped code. In \cite{WellTypedBlamed},
the authors precisely a system integrating gradual typing with contracts
\textit{à la Findler \& Felleisen}. Nickel adopts a similar type system, with
both statically typed terms, dynamically typed terms, and first-class contracts.
In this setting, each type constructor of the static type system - arrows,
records, foralls - must have a contract counterpart that is able to check at
run-time that a value is indeed a member of the corresponding type.

@@

Unions are fundamental and ubiquitous building blocks of program data. Products
represent conjunctions, that is several pieces of data whith potentially
different shapes that are packed together in one value. Unions represent
disjunctions, which are several exclusive alternatives stored in one value. When
nested together, unions and intersections are able to represent aribrary
tree-like structures, allowing to model a large class of data.

Unions are also found in configuration, to represent nullable value in JSON,
plain enumerations, or more general union types when several representations are
supported.

In statically typed languages, and more specifically in functional programming
languages, unions are usually embodied by algebraic data types [references],
also called \textbf{tagged unions}. Tagged means that there is an additional
field, or constructor, that allows to discriminate between each alternative.
When an union is constructed, the programmer specifies which alternative is
chosen by using the corresponding tag. When a union is consumed, usually via
pattern matching, code branches are tagged as well, and each branch is ensured
to handle data of the right shape. Well-known programming languages with
first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
Swift and surely many more.

\begin{lstlisting}[caption={Example of a binary tree as a tagged union in OCaml}, captionpos=b]
type binary_tree =
  | Node of 'a binary_tree * 'a binary_tree
  | Leaf of int

let sum =
  let rec aux total = function
  | Leaf val -> val
  | Node(left,right) =>
    let left = aux total left in
    aux total right
\end{lstlisting}

@@

In the statically typed setting, the actual representation of a tagged union is
an implementation detail: from the point of view of the programmer, a tagged
union is a fist-class data type associated with constructors and destructors all
with well-defined semantics. In contrast, in a gradually typed language like
Nickel, the question of the representation of tagged unions as untyped values
arises.

Unfortunately, it seems that tagged unions do not have a canonical counterpart
as serializable untyped data. A natural representation is a record with a tag
and a value which shape depends on the tag:

\begin{lstlisting}
{ tag: <A, B, C>, value: Dyn }
\end{lstlisting}

\lstinline+<A, B, C>+ is the type of a C-like enumeration that can take value
$A$, $B$ or $C$, while \lstinline+Dyn+ is the unknown dynamic type.

A better way to describe it would be as a dependent pair which first component
is a tag and second component is the data type corresponding to this tag, but
the language does not feature dependent types:
\[
    \Sigma_{tag : <A, B, C>} \text{Alternative}(\text{tag})
\]

Alas, choosing the first representation as a default is not fully satisfying.
First, it does not map precisely to a JSON type: in consequence, a lot of
illegal combinations are structurally valid as untyped JSON values. But this is
to be expected of any representation. More importantly, this does not
necessarily match with the configuration schemas one can find in the wild, that
favor simple, untagged union.

>>>>

The explorative work presented in this paper is motived by the development of
the Nickel language\cite{NickelRepo}. Nickel is a configuration programming
language, meaning that a Nickel program evaluates to a value that must be
straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
lightweight, easy to write and easy to embed in a larger system, to serve either
directly as a native language to specify configuration, or as a meta-language to
generate them. The focus is first and foremost practical: Nickel originated as a
rethink of the language of the Nix package manager, used to describe one of the
biggest sets of software packages\cite{repology}.  Nickel targets build systems
and cloud deployment tools (Kubernetes, Terraform or NixOps) as well. While
there's no hesitation in incorporating recent research when justified, the
design decisions are guided by the industrial use cases.

@@

Our aim is that writing simple Nickel feels as easy as writing JSON or YAML.
Unlike JSON, though, Nickel anticipates large configurations by being both
programmable and typed. In the configuration setting, there is a singular
dilemma with respect to typing: since a program is run on fixed inputs and is
expected to terminate, any relevant type error will show up at evaluation.  Why
bother with the complexity of a static type system? On the other hand, more and
more software systems offload complexity to configurations, such as the
infrastructure as code paradigm prevailing today in cloud deployment. When the
complexity of a codebase grows, static types becomes attractive again.  For
library code — that is, library functions —, static types are specifically
adapated and bring all the usual benefits.

This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}
which mixes both static and dynamic typing. Unlike gradually-typed languages
like TypeScript, we do not seek to build a complex type system that tries hard
to accept most of the idioms naturally arising in dynamically typed code.
Rather, we choose to provide a reasonably expressive type system with good
inference properties, that make typing generic functions operating on data easy,
but may require using untyped code for more exotic expressions.  In any case,
this is not a surrender, as Nickel provides a complementary mechanism for more
advanced data validation: contracts.

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is crucial to good error reporting. It turns out these contracts are closely
related to the type casts introduced by gradual typing, modulo blame: both
\cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the value of contracts
as a safe interface between typed and untyped code. In \cite{WellTypedBlamed},
the authors precisely introduce a system integrating gradual typing with contracts
\textit{à la Findler \& Felleisen}. Nickel adopts a similar type system, with
both statically typed terms, dynamically typed terms, and first-class contracts.
In this setting, each type constructor of the static type system - arrows,
records, foralls - must have a contract counterpart that is able to check at
run-time that a value is indeed a member of the corresponding type.

@@

Products and unions are fundamental and ubiquitous building blocks of program
data. A product is a compound value that hold several pieces of potentially
different types packed together in one value. Dually, a union only holds one
data that may alternatives stored in one value. When nested together, unions and
intersections are able to represent arbitrary tree-like structures, allowing to
model a large class of data.

Unions are also useful for configurations, to represent nullable values in JSON,
plain enumerations, or to support several representations for a field (a file
path could be accepted either as one string of "/"-separated strings, or as a
list of strings).

In statically typed languages, and more specifically in functional programming
languages, unions are usually embodied by algebraic data types [references],
also called \textbf{tagged unions}. Well-known programming languages with
first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
Swift and surely many more.

@@

In the statically typed setting, the actual representation of a tagged union is
an implementation detail: from the point of view of the programmer, a tagged
union is a fist-class data type associated with constructors and destructors,
together with well-defined semantics. In contrast, in a gradually typed language
like Nickel, the question of the representation of tagged unions as untyped
values arises.

Unfortunately, tagged unions do not have a serializable canonical counterpart as
untyped data. One usual representation is a record with a tag and a value which
shape depends on the tag (see Figure \ref{fig:union-encoding}).

\begin{figure}
  \begin{center}
    \begin{lstlisting}
        { tag: <Integer, String>, value: Dyn }
    \end{lstlisting}
  \end{center}
\caption{Encoding of the union type: integer or a string}
\label{fig:union-encoding}
\medskip
\small
In Nickel, \lstinline+<A, B, C>+ is the type of a C-like enumeration that can
take value \lstinline+A+, \lstinline+B+ or \lstinline+C+, while \lstinline+Dyn+
is the dynamic unitype.
\end{figure}

A better way to describe it would be as a dependent pair which first component
is a tag and second component is the data type corresponding to this tag:

\[
    \Sigma_{tag : <A, B, C>} \text{Alternative}(\text{tag})
\]

Alas, Nickel does not feature dependent types, which would open a whole
different level of complexity. Choosing the first representation as a default is
not fully satisfying.  First, it does not map precisely to a JSON value: in
consequence, a lot of illegal combinations are structurally valid as untyped
JSON values. But this is to be expected of any representation. More importantly,
this does not necessarily match with the configuration schemas one can find in
the wild, that favor simple, untagged union.
